---
title: "Import and QC of Zheng data set"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r load-packages}
suppressPackageStartupMessages({
  library(MultiAssayExperiment)
  library(SingleCellExperiment)
  library(scater)
  library(dplyr)
  library(tidyr)
  library(plyr)
  library(scran)
  library(cowplot)
  library(tibble)
  library(cluster)
})
```

## Load raw data

```{r load-dataset}
b.cells <- read10XResults("../../data/data_raw/zheng/b_cells/filtered_matrices_mex/hg19") 
naive.cytotoxic <- read10XResults("../../data/data_raw/zheng/naive_cytotoxic/filtered_matrices_mex/hg19")
cd14.monocytes <- read10XResults("../../data/data_raw/zheng/cd14_monocytes/filtered_matrices_mex/hg19")
regulatory.t <- read10XResults("../../data/data_raw/zheng/regulatory/filtered_matrices_mex/hg19")
cd4.t.helper <- read10XResults("../../data/data_raw/zheng/cd4_t_helper/filtered_matrices_mex/hg19")
cd56.nk <- read10XResults("../../data/data_raw/zheng/cd56_nk/filtered_matrices_mex/hg19")
memory.t <- read10XResults("../../data/data_raw/zheng/memory_t/filtered_matrices_mex/hg19")
naive.t <- read10XResults("../../data/data_raw/zheng/naive_t/filtered_matrices_mex/hg19")

colnames(b.cells) <-  paste0("b.cells", seq_len(ncol(b.cells)))
colnames(naive.cytotoxic) <- paste0("naive.cytotoxic", seq_len(ncol(naive.cytotoxic)))
colnames(cd14.monocytes) <- paste0("cd14.monocytes", seq_len(ncol(cd14.monocytes)))
colnames(regulatory.t) <- paste0("regulatory.t", seq_len(ncol(regulatory.t)))
colnames(cd4.t.helper) <- paste0("cd4.t.helper", seq_len(ncol(cd4.t.helper)))
colnames(cd56.nk) <- paste0("cd56.nk", seq_len(ncol(cd56.nk)))
colnames(memory.t) <- paste0("memory.t", seq_len(ncol(memory.t)))
colnames(naive.t) <- paste0("naive.t", seq_len(ncol(naive.t)))

colData(b.cells)$phenoid <- "b.cells"
colData(naive.cytotoxic)$phenoid <- "naive.cytotoxic"
colData(cd14.monocytes)$phenoid <- "cd14.monocytes"
colData(regulatory.t)$phenoid <- "regulatory.t"
colData(cd4.t.helper)$phenoid <- "cd4.t.helper"
colData(cd56.nk)$phenoid <- "cd56.nk"
colData(memory.t)$phenoid <- "memory.t"
colData(naive.t)$phenoid <- "naive.t"

## Subsample
set.seed(1234)
b.cells <- b.cells[, sample(colnames(b.cells), 200, replace = FALSE)]
naive.cytotoxic <- naive.cytotoxic[, sample(colnames(naive.cytotoxic), 200, replace = FALSE)]
cd14.monocytes <- cd14.monocytes[, sample(colnames(cd14.monocytes), 200, replace = FALSE)]
regulatory.t <- regulatory.t[, sample(colnames(regulatory.t), 200, replace = FALSE)]
cd4.t.helper <- cd4.t.helper[, sample(colnames(cd4.t.helper), 200, replace = FALSE)]
cd56.nk <- cd56.nk[, sample(colnames(cd56.nk), 200, replace = FALSE)]
memory.t <- memory.t[, sample(colnames(memory.t), 200, replace = FALSE)]
naive.t <- naive.t[, sample(colnames(naive.t), 200, replace = FALSE)]
```

## Create a SingleCellExperiment object

```{r create-sce}
sce <- cbind(b.cells, naive.cytotoxic, cd14.monocytes, regulatory.t, 
             cd4.t.helper, cd56.nk, memory.t, naive.t)
counts(sce) <-  as.matrix(counts(sce))
logcounts(sce) <- log2(calculateCPM(sce) + 1)
```

Exclude features that are not expressed

```{r reduce-expression-matrix}
keep_features <- rowSums(counts(sce) > 0) > 0
table(keep_features)
sce <- sce[keep_features, ]
dim(sce)
```

## Identify the remaining ERCC spike-ins.

```{r ercc-mt}
is.spike <- grepl("^ERCC", rowData(sce)$symbol)
table(is.spike)
summary(colSums(counts(sce[is.spike, ])))
isSpike(sce, "ERCC") <- which(is.spike)
```

## Calculate QC metrics

```{r QC}
sce <- calculateQCMetrics(sce, feature_controls = list(ERCC = grepl("^ERCC", rownames(sce))))
```

## Quality control on cells

```{r histogram}
size.med <- median(log10(sce$total_counts), na.rm = TRUE)
size.mad <- mad(log10(sce$total_counts), center = size.med, na.rm = TRUE)
lower.limit.size <- 10^(size.med - 3 * size.mad)
    
feature.med <- median(log10(sce$total_features), na.rm = TRUE)
feature.mad <- mad(log10(sce$total_features), center = feature.med, na.rm = TRUE)
lower.limit.feat <- 10^(feature.med - 3 * feature.mad)

ggplot(as.data.frame(colData(sce)), aes(x = total_counts)) + 
  geom_histogram(bins = 20, fill = "grey80") + xlab("Total count") + 
  ylab("Number of cells") + 
  geom_vline(xintercept = lower.limit.size, color = "red", linetype = "dashed") + 
  theme_bw()

ggplot(as.data.frame(colData(sce)), aes(x = total_features)) + 
  geom_histogram(bins = 20, fill = "grey80") + xlab("Number of detected features") + 
  ylab("Number of cells") + 
  geom_vline(xintercept = lower.limit.feat, color = "red", linetype = "dashed") + 
  theme_bw()
```

## Quality control using PCA on column data

We create a PCA plot based the quality metrics for each cell, e.g., the total
number of reads, the total number of features and the proportion of spike-in
reads.

```{r qc-pca}
sce <- scater::runPCA(sce, pca_data_input = "coldata")
scater::plotPCA(sce, colour_by = "phenoid")
```

## Filter dataset

We remove cells with log-library sizes (or total features) that are more than 3
median absolute deviations (MADs) below the median log-library size (or total
features).

```{r drop-qc}
libsize.drop <- isOutlier(sce$total_counts, nmads = 3, type = "lower", log = TRUE)
feature.drop <- isOutlier(sce$total_features, nmads = 3, type = "lower", log = TRUE)
table(libsize.drop, feature.drop)
```

```{r filter-debris}
sce <- sce[, !(libsize.drop | feature.drop)]

data.frame(ByLibSize = sum(libsize.drop), ByFeature = sum(feature.drop),
           Remaining = ncol(sce))
```

## Quality control using highest expressed genes

```{r qc-filt}
plotQC(sce, type = "highest-expression", n = 50)
```

## Gene filtering

```{r gene-filtering}
ave.counts <- rowMeans(counts(sce))
keep <- ave.counts >= 0.01
sum(keep)
df <- data.frame(ave.counts = log10(ave.counts))

ggplot(df, aes(x = ave.counts)) + 
  geom_histogram(binwidth = 0.25, fill = "grey80") + 
  geom_vline(aes(xintercept = log10(1)), color = "red", linetype = "dashed") +
  guides(color = "none") + labs(x = expression(log[10]~"(average count)"), 
                                y = "Number of genes")

sce.red <- sce[keep, ]
```

## Data normalization

```{r sizefactors}
sce <- computeSumFactors(sce, subset.row = calcAverage(sce) >= 0.001)
summary(sizeFactors(sce))

sce.red <- computeSumFactors(sce.red, subset.row = calcAverage(sce.red) >= 0.001)
summary(sizeFactors(sce.red))

ggplot(data.frame(tc = sce$total_counts, 
                  sf = sizeFactors(sce)), aes(x = sf, y = tc)) + 
  geom_point() + xlab("Size factor") + ylab("Library size") + 
  scale_x_log10() + scale_y_log10()

ggplot(data.frame(tc = sce.red$total_counts, 
                  sf = sizeFactors(sce.red)), aes(x = sf, y = tc)) + 
  geom_point() + xlab("Size factor") + ylab("Library size") + 
  scale_x_log10() + scale_y_log10()
```

```{r normalization}
sce <- normalise(sce, exprs_values = "counts", return_log = TRUE, 
                 return_norm_as_exprs = TRUE)
logcounts(sce) <- log2(calculateCPM(sce) + 1)
sce <- normalise(sce, exprs_values = "counts", return_log = FALSE, 
                 return_norm_as_exprs = FALSE)

sce.red <- normalise(sce.red, exprs_values = "counts", return_log = TRUE, 
                 return_norm_as_exprs = TRUE)
logcounts(sce.red) <- log2(calculateCPM(sce.red) + 1)
sce.red <- normalise(sce.red, exprs_values = "counts", return_log = FALSE, 
                     return_norm_as_exprs = FALSE)
```

## Plot the proportion of explained variances

```{r explained-variance, warning = FALSE} 
expl_vars <- c("phenoid", "log10_total_counts", "log10_total_features", "pct_dropout",
               "pct_counts_top_200_features", "log10_counts_feature_controls",
               "pct_counts_feature_controls")
plotQC(sce.red, type = "explanatory-variables", variables = expl_vars)
```

## Plot t-SNE representations

```{r tSNE}
set.seed(1234)
sce.red <- runTSNE(sce.red, exprs_values = "logcounts", perplexity = 5)
plotTSNE(sce.red, colour_by = "phenoid")
plotTSNE(sce.red, colour_by = "total_features", size_by = "total_counts")
```

## Save the normalized and cell filtered datasets

```{r save-data}
sce <- sce[!isSpike(sce, "ERCC"), ]
sce.red <- sce.red[!isSpike(sce.red, "ERCC"), ]
saveRDS(sce.red, file = "../../data/sce_filteredExpr/sce_filteredExpr_Zhengmix8eq.rds")
saveRDS(sce, file = "../../data/sce_full/sce_full_Zhengmix8eq.rds")
```

## Session info

```{r}
date()
sessionInfo()
```

