---
title: "Import and QC of Kumar data set (GSE60749-GPL13112)"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r load-packages}
suppressPackageStartupMessages({
  library(MultiAssayExperiment)
  library(SingleCellExperiment)
  library(scater)
  library(scran)
  library(plyr)
  library(dplyr)
  library(ggplot2)
})
```

## Load MultiAssayExperiment object

```{r load-dataset}
(maex <- readRDS("../../data/data_raw/GSE60749-GPL13112.rds"))

## Extract the gene-level length-scaled TPMs
cts <- assays(experiments(maex)[["gene"]])[["count_lstpm"]]

## Extract the phenotype data.
phn <- colData(maex)
phn$phenoid <- as.character(interaction(as.data.frame(phn[, c("source_name_ch1",
                                                              "characteristics_ch1.1")])))
## Simplify labels
phn$phenoid <-  plyr::revalue(
  phn$phenoid, 
  c("Dgcr8 knockout mouse embryonic stem cells.culture conditions: serum+LIF" = "Dgcr8 knockout mouse serum+LIF", 
    "v6.5 mouse embryonic stem cells.culture conditions: 2i+LIF" = "v6.5 mouse 2i+LIF",
    "v6.5 mouse embryonic stem cells.culture conditions: serum+LIF" = "v6.5 mouse serum+LIF")
)
table(phn$phenoid)
```

## Create a SingleCellExperiment object

```{r create-sce}
stopifnot(all(colnames(cts) == rownames(phn)))

sce <- SingleCellExperiment(
  assays = list(counts = cts), 
  colData = phn
)
sce <- normalise(sce, exprs_values = "counts", return_log = TRUE, 
                 return_norm_as_exprs = TRUE) ## generates logcounts(sce)
```

Exclude features that are not expressed

```{r reduce-expression-matrix}
keep_features <- rowSums(counts(sce) > 0) > 0
table(keep_features)
sce <- sce[keep_features, ]
dim(sce)
```

## Identify the remaining ERCC spike-ins.

```{r ercc-mt}
is.spike <- grepl("^ERCC", rownames(sce))
table(is.spike)
summary(colSums(counts(sce[is.spike, ])))
isSpike(sce, "ERCC") <- is.spike
```

## Calculate QC metrics

```{r QC}
sce <- calculateQCMetrics(sce, feature_controls = list(ERCC = grepl("^ERCC", rownames(sce))))
```

## Quality control using PCA on column data

We create a PCA plot based the quality metrics for each cell, e.g., the total
number of reads, the total number of features and the proportion of spike-in
reads.

```{r qc-pca}
sce <- scater::runPCA(sce, pca_data_input = "coldata")
scater::plotPCA(sce, colour_by = "phenoid")
```

## Filter cells

We remove cells with log-library sizes (or total features) that are more than 3
median absolute deviations (MADs) below the median log-library size (or total
features).

```{r histogram}
colData(sce)$libsize.drop <- isOutlier(sce$total_counts, nmads = 3, type = "lower", log = TRUE)
ggplot(as.data.frame(colData(sce)), aes(x = total_counts)) + 
  geom_histogram(bins = 20, fill = "grey80") + xlab("Total count") + 
  ylab("Number of cells") + 
  geom_vline(xintercept = min(sce$total_counts[!sce$libsize.drop]), 
             color = "red", linetype = "dashed") + 
  theme_bw()

colData(sce)$feature.drop <- isOutlier(sce$total_features, nmads = 3, type = "lower", log = TRUE)
ggplot(as.data.frame(colData(sce)), aes(x = total_features)) + 
  geom_histogram(bins = 20, fill = "grey80") + xlab("Number of detected features") + 
  ylab("Number of cells") + 
  geom_vline(xintercept = min(sce$total_features[!sce$feature.drop]), 
             color = "red", linetype = "dashed") + 
  theme_bw()

table(libsize = sce$libsize.drop, feature = sce$feature.drop)
```

We also filter out cells with a large fraction of ERCC reads.

```{r filter-ercc}
colData(sce)$spike.drop <- isOutlier(sce$pct_counts_ERCC, nmads = 3, type = "higher")
ggplot(as.data.frame(colData(sce)), aes(x = pct_counts_ERCC)) + 
  geom_histogram(bins = 20, fill = "grey80") + xlab("ERCC proportion (%)") + 
  ylab("Number of cells") + 
  geom_vline(xintercept = max(sce$pct_counts_ERCC[!sce$spike.drop]), 
             color = "red", linetype = "dashed") + 
  theme_bw()

table(sce$spike.drop)
sce <- sce[, !(sce$libsize.drop | sce$feature.drop | sce$spike.drop)]
dim(sce)
```

## Quality control using highest expressed genes

```{r qc-filt}
plotQC(sce, type = "highest-expression", n = 50)
```

## Data normalization

```{r sizefactors}
sce <- computeSumFactors(sce, sizes = pmin(ncol(sce), seq(20, 120, 20)), min.mean = 0.1)
summary(sizeFactors(sce))
sce <- computeSpikeFactors(sce, general.use = FALSE)
```

```{r normalization}
sce <- normalise(sce, exprs_values = "counts", return_log = TRUE, 
                 return_norm_as_exprs = TRUE)
sce <- normalise(sce, exprs_values = "counts", return_log = FALSE, 
                 return_norm_as_exprs = FALSE)
```

## Plot the proportion of explained variances

```{r explained-variance, warning = FALSE} 
expl_vars <- c("phenoid", "log10_total_counts", "log10_total_features", "pct_dropout",
               "pct_counts_top_200_features", "log10_counts_feature_controls",
               "pct_counts_feature_controls")
plotQC(sce, type = "explanatory-variables", variables = expl_vars)
```

## Plot t-SNE representations

```{r tSNE}
set.seed(1234)
sce <- runTSNE(sce, exprs_values = "logcounts", perplexity = 10)
plotTSNE(sce, colour_by = "phenoid")
plotTSNE(sce, colour_by = "total_features", size_by = "total_counts")
```

## Save the normalized and cell filtered dataset

```{r save-data}
sce <- sce[!isSpike(sce, "ERCC"), ]
dim(sce)
table(sce$phenoid)
saveRDS(sce, file = "../../data/sce_full/sce_full_Kumar.rds")
```

## Session info

```{r}
date()
sessionInfo()
```

