---
title: "Import and QC of Koh data set"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Load and process data set SRP073808 (Koh 2016)

```{r load-packages}
suppressPackageStartupMessages({
  library(MultiAssayExperiment)
  library(scater)
  library(dplyr)
  library(tidyr)
  library(plyr)
  library(scran)
  library(cowplot)
  library(tibble)
})
```

## Load MultiAssayExperiment object

```{r load-dataset}
(maex <- readRDS("../../data/data_raw/SRP073808.rds"))

## Extract the gene-level TPM and the counts obtaines by scaling the TPMs
cts <- assays(experiments(maex)[["gene"]])[["count_lstpm"]]
tpms <- assays(experiments(maex)[["gene"]])[["TPM"]]

## Extract the phenotype data. 
phn <- colData(maex)
phn$phenoid <- as.character(phn$LibraryName)

table(phn$phenoid)
```

## Create a SingleCellExperiment object

```{r create-sce}
sce <- SingleCellExperiment(
  assays = list(counts = cts, 
                tpm = tpms), 
  colData = phn)
logcounts(sce) <- log2(calculateCPM(sce) + 1)
```

Exclude features that are not expressed

```{r reduce-expression-matrix}
keep_features <- rowSums(counts(sce) > 0) >= 1
table(keep_features)
sce <- sce[keep_features, ]
dim(sce)
```

## Identify the ERCC spike-ins.

There are 92 ERCC spike-in genes, but no mitochondrial genes.

```{r ercc-mt}
is.spike <- grepl("^ERCC", rownames(sce))
is.mito <- grepl("^mt-", rownames(sce))
table(is.spike)
table(is.mito)
```

## Calculate QC metrics

```{r QC}
sce <- calculateQCMetrics(sce)
```

## Quality control on cells

```{r histogram}
size.med <- median(log10(sce$total_counts), na.rm = TRUE)
size.mad <- mad(log10(sce$total_counts), center = size.med, na.rm = TRUE)
lower.limit.size <- 10^(size.med - 3 * size.mad)/1e6
    
feature.med <- median(log10(sce$total_features), na.rm = TRUE)
feature.mad <- mad(log10(sce$total_features), center = size.med, na.rm = TRUE)
lower.limit.feat <- 10^(feature.med - 3 * feature.mad)

ggplot(as.data.frame(colData(sce)), aes(x = total_counts/1e6)) + 
  geom_histogram(bins = 20, fill = "grey80") + xlab("Library size (million reads)") + 
  ylab("Number of cells") + 
  geom_vline(xintercept = lower.limit.size, color = "red", linetype = "dashed") + 
  theme_bw()

ggplot(as.data.frame(colData(sce)), aes(x = total_features)) + 
  geom_histogram(bins = 20, fill = "grey80") + xlab("Number of detected features") + 
  ylab("Number of cells") + 
  geom_vline(xintercept = lower.limit.feat, color = "red", linetype = "dashed") + 
  theme_bw()
```

## Quality control using PCA and highest expressed genes

We create a PCA plot based the quality metrics for each cell, e.g., the total
number of reads, the total number of features and the proportion of
mitochondrial or spike-in reads.

```{r qc-pca}
p1 <- scater::plotPCA(sce, exprs_values = "logcounts",
                      selected_variables = "pdata",colour_by = "phenoid") + 
  theme(legend.position = "bottom") + 
  guides(color = guide_legend(nrow = 3, byrow = TRUE))
p2 <- scater::plotQC(sce, type = "find-pcs", variable = "total_features")
```

## Filter dataset

We remove cells with log-library sizes that are more than 3 median absolute
deviations (MADs) below the median log-library size.

```{r drop-qc}
libsize.drop <- isOutlier(sce$total_counts, nmads = 3, type = "lower", log = TRUE)
feature.drop <- isOutlier(sce$total_features, nmads = 3, type = "lower", log = TRUE)
table(libsize.drop, feature.drop)
```

```{r filter-debris}
sce <- sce[, !(libsize.drop | feature.drop)]

data.frame(ByLibSize = sum(libsize.drop), ByFeature = sum(feature.drop),
           Remaining = ncol(sce))
```

## Quality control using PCA and highest expressed genes

Using the filtered data, we make a PCA plot based the quality metrics for each
cell, e.g., the total number of reads, the total number of features and the
proportion of mitochondrial or spike-in reads.

```{r qc-filt}
p3 <- plotQC(sce, type = "highest-expression", n = 50)
```

```{r mean-freq-expr}
plotQC(sce, type = "exprs-freq-vs-mean")
```

## Gene filtering

```{r gene-filtering}
ave.counts <- rowMeans(counts(sce))
keep <- ave.counts >= 1
sum(keep)
df <- data.frame(ave.counts = log10(ave.counts))

p4 <- ggplot(df, aes(x = ave.counts)) + 
  geom_histogram(binwidth = 0.5) + 
  geom_vline(aes(xintercept = log10(1), color = "red")) +
  guides(color = "none") + labs(x = expression(Log[10]~"average count"), 
                                y = "Number of genes")

sce.red <- sce[keep, ]
```

## Data normalization

```{r sizefactors}
sce <- computeSumFactors(sce, sizes = c(20, 40, 80, 150))
summary(sizeFactors(sce))

sce.red <- computeSumFactors(sce.red, sizes = c(20, 40, 80, 150))
summary(sizeFactors(sce.red))

ggplot(data.frame(tc = sce$total_counts/1e6, 
                  sf = sizeFactors(sce)), aes(x = sf, y = tc)) + 
  geom_point() + xlab("Size factor") + ylab("Library size (million reads)") + 
  scale_x_log10() + scale_y_log10()

ggplot(data.frame(tc = sce.red$total_counts/1e6, 
                  sf = sizeFactors(sce.red)), aes(x = sf, y = tc)) + 
  geom_point() + xlab("Size factor") + ylab("Library size (million reads)") + 
  scale_x_log10() + scale_y_log10()
```

```{r normalization}
sce <- normalise(sce, exprs_values = "counts", return_log = TRUE, 
                 return_norm_as_exprs = TRUE))
logcounts(sce) <- log2(calculateCPM(sce) + 1)
sce <- normalise(sce, exprs_values = "counts", return_log = FALSE, 
                 return_norm_as_exprs = FALSE)

sce.red <- normalise(sce.red, exprs_values = "counts", return_log = TRUE, 
                 return_norm_as_exprs = TRUE))
logcounts(sce.red) <- log2(calculateCPM(sce.red) + 1)
sce.red <- normalise(sce.red, exprs_values = "counts", return_log = FALSE, 
                     return_norm_as_exprs = FALSE)
```

## Plot the proportion of explained variances

```{r explained-variance} 
expl_vars <- c("phenoid", "log10_total_counts", "log10_total_features", "pct_dropout",
               "pct_counts_top_200_features", "log10_counts_feature_controls",
               "pct_counts_feature_controls")
p5 <- plotQC(sce.red, type = "explanatory-variables", variables = expl_vars)
```

## Plot t-SNE representations

```{r tSNE}
set.seed(1234)
p6 <- plotTSNE(sce.red, exprs_values = "logcounts", perplexity = 5,
               colour_by = "total_features", size_by = "total_counts",
               return_SCESet=FALSE) + ggtitle("tSNE plot")
```

## Plot a summary of the QC

```{r qc-plot-out} 
plot.qc <- plot_grid(p1, p2, p3, p4, p5, p6, ncol = 2, labels = "auto")
pdf("../../plots/qc_data/qc_summary_Koh.pdf", width = 10, height = 12)
print(plot.qc)
dev.off()
```

## Save the normalized and cell filtered datasets

```{r save-data}
saveRDS(sce, file = "../../data/sce_full/sce_full_Koh.rds")
saveRDS(sce.red, file = "../../data/sce_filtered/sce_filtered_Koh.rds")
```

## Session info

```{r}
date()
sessionInfo()
```

