---
title: "Import and QC of SimKumar data set"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r load-packages}
suppressPackageStartupMessages({
  library(MultiAssayExperiment)
  library(SingleCellExperiment)
  library(scater)
  library(scran)
  library(dplyr)
  library(ggplot2)
  library(splatter)
})
```

## Load MultiAssayExperiment object

```{r load-dataset}
(maex <- readRDS("../../data/data_raw/GSE60749-GPL13112.rds"))

## Extract the gene-level length-scaled TPMs
cts <- assays(experiments(maex)[["gene"]])[["count_lstpm"]]

## Extract the phenotype data. 
phn <- colData(maex)
phn$phenoid <- as.character(interaction(as.data.frame(phn[, c("source_name_ch1",
                                                              "characteristics_ch1.1")])))
## Simplify labels
phn$phenoid <-  plyr::revalue(
  phn$phenoid , 
  c("Dgcr8 knockout mouse embryonic stem cells.culture conditions: serum+LIF" = "Dgcr8 knockout mouse serum+LIF", 
    "v6.5 mouse embryonic stem cells.culture conditions: 2i+LIF" = "v6.5 mouse 2i+LIF",
    "v6.5 mouse embryonic stem cells.culture conditions: serum+LIF" = "v6.5 mouse serum+LIF")
)
table(phn$phenoid)
```

## Extract one subpopulation

```{r}
stopifnot(all(colnames(cts) == rownames(phn)))

sub.cts <- cts[, phn$phenoid == "v6.5 mouse 2i+LIF"]
dim(sub.cts)
```

## Remove the ERCC spike-ins

```{r remove-spikes}
is.spike <- grepl("^ERCC", rownames(sub.cts))
sub.cts <- sub.cts[!is.spike, ]
```

## Estimate parameters for simulation

```{r estimate-parameters}
(params <- splatEstimate(sub.cts))
```

## Simulate new counts

```{r simulate-counts}
sce <- splatSimulate(params = params, batchCells = 500,
                     group.prob = c(0.13, 0.07, 0.1, 0.05, 0.4, 0.1, 0.1, 0.05), 
                     method = "groups", 
                     de.prob = c(0.01, 0.03, 0.03, 0.05, 0.05, 0.07, 0.08, 0.1),
                     verbose = FALSE)
sce <- normalise(sce, exprs_values = "counts", return_log = TRUE, 
                 return_norm_as_exprs = TRUE) ## generates logcounts(sce)
sce$phenoid <- colData(sce)$Group
```

Exclude features that are not expressed

```{r reduce-expression-matrix}
keep_features <- rowSums(counts(sce) > 0) > 0
table(keep_features)
sce <- sce[keep_features, ]
dim(sce)
```

## Calculate QC metrics

```{r QC}
sce <- calculateQCMetrics(sce)
```

## Quality control using PCA on column data

We create a PCA plot based the quality metrics for each cell, e.g., the total
number of reads and the total number of features.

```{r qc-pca}
sce <- scater::runPCA(sce, pca_data_input = "coldata")
scater::plotPCA(sce, colour_by = "phenoid")
```

## Filter cells

We remove cells with log-library sizes (or total features) that are more than 3
median absolute deviations (MADs) below the median log-library size (or total
features).

```{r histogram}
colData(sce)$libsize.drop <- isOutlier(sce$total_counts, nmads = 3, type = "lower", log = TRUE)
ggplot(as.data.frame(colData(sce)), aes(x = total_counts)) + 
  geom_histogram(bins = 20, fill = "grey80") + xlab("Total count") + 
  ylab("Number of cells") + 
  geom_vline(xintercept = min(sce$total_counts[!sce$libsize.drop]), 
             color = "red", linetype = "dashed") + 
  theme_bw()

colData(sce)$feature.drop <- isOutlier(sce$total_features, nmads = 3, type = "lower", log = TRUE)
ggplot(as.data.frame(colData(sce)), aes(x = total_features)) + 
  geom_histogram(bins = 20, fill = "grey80") + xlab("Number of detected features") + 
  ylab("Number of cells") + 
  geom_vline(xintercept = min(sce$total_features[!sce$feature.drop]), 
             color = "red", linetype = "dashed") + 
  theme_bw()
```

```{r filter-cells}
table(sce$libsize.drop, sce$feature.drop)
sce <- sce[, !(sce$libsize.drop | sce$feature.drop)]
dim(sce)
```

## Quality control using highest expressed genes

```{r qc-filt}
plotQC(sce, type = "highest-expression", n = 50)
```

## Gene filtering

```{r gene-filtering}
ave.counts <- rowMeans(counts(sce))
keep <- ave.counts >= 1
sum(keep)
df <- data.frame(ave.counts = log10(ave.counts))

ggplot(df, aes(x = ave.counts)) + 
  geom_histogram(binwidth = 0.25, fill = "grey80") + 
  geom_vline(aes(xintercept = log10(1), color = "red")) +
  guides(color = "none") + labs(x = expression(log[10]~"(average count)"), 
                                y = "Number of genes")

sce.red <- sce[keep, ]
```

## Data normalization

```{r sizefactors}
sce <- computeSumFactors(sce, sizes = c(20, 40, 60, 80))
summary(sizeFactors(sce))

sce.red <- computeSumFactors(sce.red, sizes = c(20, 40, 60, 80))
summary(sizeFactors(sce.red))
```

```{r normalization}
sce <- normalise(sce, exprs_values = "counts", return_log = TRUE, 
                 return_norm_as_exprs = TRUE)
sce <- normalise(sce, exprs_values = "counts", return_log = FALSE, 
                 return_norm_as_exprs = FALSE)

sce.red <- normalise(sce.red, exprs_values = "counts", return_log = TRUE, 
                 return_norm_as_exprs = TRUE)
sce.red <- normalise(sce.red, exprs_values = "counts", return_log = FALSE, 
                     return_norm_as_exprs = FALSE)
```

## Plot the proportion of explained variances

```{r explained-variance, warning = FALSE} 
expl_vars <- c("phenoid", "log10_total_counts", "log10_total_features", "pct_dropout",
               "pct_counts_top_200_features", "log10_counts_feature_controls",
               "pct_counts_feature_controls")
plotQC(sce.red, type = "explanatory-variables", variables = expl_vars)
```

## Plot t-SNE representations

```{r tSNE}
set.seed(1234)
sce.red <- runTSNE(sce.red, exprs_values = "logcounts", perplexity = 5)
plotTSNE(sce.red, colour_by = "phenoid")
plotTSNE(sce.red, colour_by = "total_features", size_by = "total_counts")
```

## Save the normalized and cell filtered datasets

```{r save-data}
saveRDS(sce.red, file = "../../data/sce_filteredExpr/sce_filteredExpr_SimKumar8hard.rds")
saveRDS(sce, file = "../../data/sce_full/sce_full_SimKumar8hard.rds")
```

## Session info

```{r}
date()
sessionInfo()
```

